Assessment of simulation procedures and robust methods for Mendelian randomization analysis

### 2022-11-08 presentation/robustMR.qmd
revealjs presentation 2022-11-08

---
title: "`r params$doc_title`"
author: | 
        | christoph.reich@med.uni-heidelberg.de 
        | lorenzo@imbi.uni-heidelberg.de
        | 
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
params:
  date: !r Sys.Date()
  doc_title: "MSc Default Title"
output: 
  html_document:
    theme: paper
    toc: true
    number_sections: true
    toc_float: true
    toc_depth: 4
    df_print: kable
    latex_engine: xelatex
#bibliography: ../references/project01.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(ggplot2)
library(scales)
library(ggthemes)
#library(biomaRt)  # different script
library(kableExtra)
library(MendelianRandomization)
library(mr.raps)
library(MRMix)
library(MRPRESSO)
library(penalized)
library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
```

```{r}
my_data <- as_tibble(read.delim("../data-boekstegers_felix/GST_BMI_association_results.txt"))
# GBC: new data 2022-11-04
GBC_dat <- as_tibble(read.delim("../data-boekstegers_felix/C23C24_BMI_association_results.txt"))

@@

# Get gene names from rs-ids

```{r}
# see script biomart_search.R
rsIds_GeneNames <- as_tibble(readRDS("../data/biomartBMIsearch.rds"))

filter_FTO <- rsIds_GeneNames %>% 
  filter(str_detect(associated_gene, pattern = "FTO")) %>% 
  distinct(refsnp_id, .keep_all = TRUE)
```

# Simulating on empirical data

## Calculate approximate SEs for decreasing sample sizes and corresponding p-vals

Calculating p-values from a normal distribution. The se, pval and study size depend on each other, within each study. The selection of IVs should rely on the exposure study - BMI in our case. Summary statistics on genetic associations with BMI will be based on results of the GIANT consortium + GERA (n=334487 (100,418 + 234,069) individuals and 289 associated variants with p < 10-8) **(insert citation)**. We vary the sample size and calculate the number of IVs and the explained variance. *(Justo Mail 2022-09-27)*
\
Steps:
1.       Calculate more accurates SEs using the betas and Pvals provided to you (this will be the original/initial SE for n=334.487). 
2.       Calculate the approximate SEs for decreasing samples sizes
3.       Calculate the probability values considering the estimated betas and the corresponding SEs
4.       Exclude the IV if Pval > genome-wide significance threshold  (5 × 10−8) **NEW 5x10^(-6))**
5.       Calculate cumulative explained variances
6.       Represent figure similar to Fig1 in Chatterjee’s paper

For the selected IVs, please also calculate the explained variance for the confounder and outcome.

```{r z-statistic}
### justo approach email 30-10-2022  ---------------------------------------

## LOOP simulation ----------------------------------------------
n.orig <- 334487
n <- c(1000, 2000, 5000, seq(10000, 330000, by=10000), 334487)

@@

# 4. Exclude the IV if Pval > genome-wide significance threshold  (5 × 10−8)
pvals_df %>% 
  summarize(across(.cols = everything(), function(x) sum(x < 5*10^(-6)))) %>% 
  t() %>% 
  kable(caption = "Count of significant SNPs depending on sample size")

# pvals_df %>%
#   filter(p_value_exp_sample_100000< 5*10^(-6)) %>% 
#   count()

my_data <- as_tibble(cbind(my_data, pvals_df))

# 5. Calculate cumulative explained variances
# 6. Represent figure similar to Fig1 in Chatterjee’s paper
```

```{r calculate-explained-variance}
explained_variance_numeric <- function(eaf, beta) {
  # eaf = allele frequency
  # beta = additive genetic effect
  explained_variance_numeric <- 2*eaf *(1-eaf) * beta^2
  return(explained_variance_numeric)
}

@@

```{r}
# labeling (only FTO label)
my_data$gene_label <- NA
my_data$gene_label[my_data$SNP %in% filter_FTO$refsnp_id] <- "FTO"

@@

@@

```{r plot-fig1}
# 6.       Represent figure similar to Fig1 in Chatterjee’s paper

@@

ggplot(plot_data, aes(x=log10(n), y=value, col=name))+
  geom_point(alpha=0.6)+
  geom_line(alpha=0.6)+
  xlab("log10(N)")+
  #ylab("% variance explained by IVs")+
  labs(title = "Figure 1",
       col='', 
       caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       ) +
  scale_y_continuous(
    # Features of the first axis
    name = "% variance explained by IVs",
    # Add a second axis and specify its features
    sec.axis = sec_axis(~./scale_factor, name="Number of IVs")
  ) + 
  theme_few()+
  scale_colour_few() ->fig01
fig01

@@

# Different genetic architectures of the exposure

## MAF vs P-val

For the ~200 IVs, please plot the minor allele frequency (x-axis) versus -log10Pval (y-axis). Calculate the median MAF and the median –log10Pval, and draw the corresponding vertical and horizontal lines. The four areas will define four genetic architectures for the simulations (e.g. relatively rare variants with relatively large exposure effects).

```{r MAF-Pval}
ggplot(data = my_data, aes(x=eaf_exp, y=-log10(p_value_exp)))+
  geom_point(alpha=0.6)+
  geom_hline(yintercept=-log10(median(my_data$p_value_exp)), col="red", alpha=0.6)+
  geom_vline(xintercept=median(my_data$eaf_exp), col="red", alpha=0.6)+
  xlab("Effect allele frequency for exposure")+
  ylab("-log10(P-value) ")+
  labs(title = "Figure 2",
       subtitle = "Define different genetic architectures",
       #col='', 
       #caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       ) +
  theme_few()+
  scale_colour_few() ->fig02
fig02

@@

## MAF vs betas

@@

# Simulation Framework

- Let’s try to estimate parameters for the simulation (pi1, varx, pi2, varu, muy, vary, pi3, pi4, delta0) based on real data. Maybe you could start with N=10.000 and N=300.000 (I will help with this)

>>>>

---
title: "Assessment of simulation procedures and robust methods for Mendelian randomization analysis"
author: "Christoph Reich"
date: "`r Sys.Date()`"
title-slide-attributes: 
  data-background-image: ../templates/bg1.png
  data-background-size: cover  # default
  data-background-opacity: "1"  # from 0-1
format: 
  revealjs:
    slide-number: true
    scrollable: true
    navigation-mode: vertical
    chalkboard:   # kreidetafel
      theme: whiteboard
      boardmarker-width: 5
      buttons: false
    incremental: false  # display bullet lists all at once
    preview-links: auto
    logo: ../templates/mederlab_logo.png
    css: ../templates/main.css
    footer: "robustMRmethods"
    theme: [simple, ../templates/custom.scss]
    #self-contained: true  # for distribution (slows rendering), not compatible with chulkboarddic
# reference-doc: templates/20210129_mederslidemaster.pptx  # only for pptx rendering
bibliography: ../references/references.bib
---

```{r}
#| label: dependencies
#| include: false
#| warning: false

rm(list=ls())

knitr::opts_chunk$set(echo = FALSE)

library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(ggplot2)
library(scales)
library(ggthemes)
#library(biomaRt)  # different script
library(kableExtra)
library(MendelianRandomization)
library(mr.raps)
library(MRMix)
library(MRPRESSO)
library(penalized)
library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")

theme_set(ggthemes::theme_few()) 
```

```{r}
#| label: load-data
#| include: false
#| warning: false

my_data <- as_tibble(read.delim("../data-boekstegers_felix/GST_BMI_association_results.txt"))
# GBC: new data 2022-11-04
GBC_dat <- as_tibble(read.delim("../data-boekstegers_felix/C23C24_BMI_association_results.txt"))

@@

## Get gene names from rs-ids

```{r}
#| label: FTO-annotation
#| tbl-cap: "SNP associated with FTO gene"
#| tbl-colwidths: [60,40]

# see script biomart_search.R
rsIds_GeneNames <- as_tibble(readRDS("../data/biomartBMIsearch.rds"))

filter_FTO <- rsIds_GeneNames %>% 
  filter(str_detect(associated_gene, pattern = "FTO")) %>% 
  distinct(refsnp_id, .keep_all = TRUE)

t(filter_FTO) %>% kable() %>% 
  kable_styling(font_size=18)
```

:::: {.columns}

::: {.column width="50%"}
* from the 284 SNPs only one was associated with the FTO gene
:::

::: {.column width="50%"}
![](../img/ensembl_logo.png)

:::

::::

## Simulating on empirical data

### Calculate approximate SEs for decreasing sample sizes and corresponding p-vals

* se, pval and study size depend on each other, within each study
* selection of IVs based on exposure study (BMI in our case)
* Summary statistics on genetic associations with BMI based on results of the GIANT + GERA studies [@Hoffmann_2018] with $N=100,418 + 234,069 =334487$  and $SNPs=289$

## Steps

1. Calculate more accurates SEs using the betas and Pvals
2. Calculate the approximate SEs for decreasing samples sizes
3. Calculate the probability values considering the estimated betas and the corresponding SEs
4. Exclude the IV if Pval > genome-wide significance threshold  (5 × 10−8) **NEW 5x10^(-6))**
5. Calculate cumulative explained variances
6. Represent figure similar to Fig1 in Chatterjee’s paper @QiChatterjee_2021

```{r}
#| label: z-statistic

### justo approach email 30-10-2022  ---------------------------------------
## LOOP simulation ----------------------------------------------
n.orig <- 334487
n <- c(1000, 2000, 5000, seq(10000, 330000, by=10000), 334487)

@@

# 4. Exclude the IV if Pval > genome-wide significance threshold  (5 × 10−8)
my_data <- as_tibble(cbind(my_data, pvals_df))

pvals_df %>% 
  summarize(across(.cols = everything(), function(x) sum(x < 5*10^(-6)))) %>% 
  t() %>% 
  kable(caption = "Count of significant SNPs depending on sample size") %>% 
  kable_styling(font_size=18)
```

## Explained variance

* The variance in BMI explained by the IVs is calculated with the following formula: $explained\ variance = 2f(1-f)\beta^2$, where $f$ denotes the allele frequency and $\beta$ is the additive genetic effect. 
* The explained variance in liability to GSD and GBC is calculated as described by @So_2011.

```{r}
#| label: calculate-explained-variance

explained_variance_numeric <- function(eaf, beta) {
  # eaf = allele frequency
  # beta = additive genetic effect
  explained_variance_numeric <- 2*eaf *(1-eaf) * beta^2
  return(explained_variance_numeric)
}

@@

```{r}
#| label: explainedBMIvarianceVSpval
#|

my_data$gene_label <- NA
my_data$gene_label[my_data$SNP %in% filter_FTO$refsnp_id] <- "FTO"

@@

Here we show a plot similar to Figure 1 in [@QiChatterjee_2021]:

@@

```{r}
#| label: figure1QiChatterjee
# 6.       Represent figure similar to Fig1 in Chatterjee’s paper

@@

ggplot(plot_data, aes(x=log10(n), y=value, col=name))+
  geom_point(alpha=0.6)+
  geom_line(alpha=0.6)+
  xlab("log10(N)")+
  #ylab("% variance explained by IVs")+
  labs(title = "Figure 1",
       col='', 
       caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       ) +
  scale_y_continuous(
    # Features of the first axis
    name = "% variance explained by IVs",
    # Add a second axis and specify its features
    sec.axis = sec_axis(~./scale_factor, name="Number of IVs")
  ) + 
  theme_few()+
  scale_colour_few() ->fig01

@@

## Different genetic architectures of the exposure

```{r MAF-Pval}
ggplot(data = my_data, aes(x=eaf_exp, y=-log10(p_value_exp)))+
  geom_point(alpha=0.6)+
  geom_hline(yintercept=-log10(median(my_data$p_value_exp)), col="red", alpha=0.6)+
  geom_vline(xintercept=median(my_data$eaf_exp), col="red", alpha=0.6)+
  xlab("Effect allele frequency for exposure")+
  ylab("-log10(P-value) ")+
  labs(title = "Figure 2",
       subtitle = "Define different genetic architectures",
       #col='', 
       #caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       ) +
  theme_few()+
  scale_colour_few() ->fig02
#fig02

@@

### MAF vs betas

@@

## Simulation Framework

* Estimate parameters for the simulation (pi1, varx, pi2, varu, muy, vary, pi3, pi4, delta0) based on real data
* Start with $N=10.000$ and $N=300.000$

![](../img/m4qi2021.png)

##

**References**

::: {#refs}
:::

**Software**

This analysis was carried out using the statistical software `r base::version$version.string` along with [RStudio](https://www.rstudio.com).
Apart from R's base functionality, the following packages were used: <br/>

```{r software, echo=FALSE}
# or faster with function...
installed.packages()[names(sessionInfo()$otherPkgs), "Version"]
```

##  {background-image="../templates/bg4.png"}

::: footer

:::
